\textbf{Varying coefficients correlated velocity models in complex landscapes with application to narwhal behavioral response to ship exposure and airgun exposure}

### 2024-06-01 Latex/main.tex
Last paper version

@@

We focus on continuous-time models defined as a solution of a stochastic differential equation with time-varying coefficients. One widely known model for such data is the continuous-time correlated random walk, also known as the integrated Ornstein-Uhlenbeck or correlated velocity model (CVM) \cite{johnson_continuoustime_2008}. It can be viewed as a special case of velocity potential model, where the potential is a second degree polynomial \cite{preisler_analyzing_2013}. The following formulation of the equations represents an extension of the CVM which involves a rotational component and  is based on biologically interpretable parameters \cite{gurarie_correlated_2017}.
\begin{equation} \left\{
	\begin{array}{l}
		dX(t)=V(t)dt \\
		dV(t)=-\begin{pmatrix} 
			\frac{1}{\tau} & -\omega \\
			\omega & \frac{1}{\tau}
		\end{pmatrix}(V(t)-\mu)dt+\frac{2\nu}{\sqrt{\pi \tau}} dW(t) 
	\end{array}
	\right.
	\label{eq: RACVM equation}
\end{equation}
where $V(t)$ is the horizontal two-dimensional velocity at time $t$, $X(t)$ is the position, typically in longitude and latitude or in UTM coordinates and $W(t)$ is a two-dimensional brownian motion. 
The parameter $\tau$ is an autocorrelation time scale, $\nu$ controls the norm of the velocity and drives the random variability, $\mu$ is the long term velocity, and $\omega$ is an angular velocity that controls how fast the velocity vector rotates. The case $\omega=0$ corresponds to a standard CVM. In most applications, this standard model is satisfactory, but in case of tortuous movement, a non-zero value of $\omega$ is sometimes needed \cite{gurarie_correlated_2017,alt_correlation_1990,albertsen_generalizing_2018}. Moreover, whether or not to include a non-zero mean velocity parameter $\mu$ depends on the specific context of the study. Typically, it would make sense to have $\mu\neq (0,0)$ when examining migratory patterns or avoidance reactions from a fixed sound source. We follow \cite{gurarie_correlated_2017} and refer to~\ref{eq: RACVM equation} as "Rotational Correlated Velocity Model" (RCVM) in the case $\mu=(0,0)$, and "Rotational Advective Correlated Velocity Model" (RACVM) in the case $\mu \neq (0,0)$.\\

In general, the motion of marine mammals is subject to spatial constraints induced by the coastline. This is particularly the case for the narwhal data we want to analyse: the animals move in a restricted domain that consists in a complex system of fjords in Scoresby Sound (South-East Greenland). Thus, constraints for the shore are needed in the location process $X$. However, this is not the case in model~\ref{eq: RACVM equation}, nor in most of the SDE used for animal movement \cite{johnson_continuoustime_2008,michelot_varying-coefficient_2021,gurarie_correlated_2017}. 
Recently, a reflected version of the CVM has been considered to study sea lion telemetry data \cite{hanks_reflected_2017}.
In this study, simulation of constrained paths is achieved by projecting on the coastline the  points that eventually reach land. Estimation of the SDE parameters then relies on computationally intensive MCMC algorithm (Metroplis-Hastings), and requires high frequency data. Another idea is to add a repulsive potential function in the drift, which is done to constrain ants movement within a box in \cite{russell_spatially_2018}. In our case study, the narwhals move in a complex fjords system, which means that this technique might not be easily replicated because it relies on using very basic boundaries to derive an explicit expression for the potential function. As a consequence, there is a need to develop new models adapted to such spatial constraints.\\
Additionally, the way to model the motion of an animal near the boundary of the domain should be species-specific, accounting for the particular behaviors and spatial use of the studied animal \cite{brillinger_simulating_2003}. For narwhals, it was observed from the data that they have a proclivity to move along the shoreline. Here we propose a new model based on the following idea: as the narwhals approach the shore, at some point they need to rotate to avoid or align with the shoreline. The novelty of this paper consists in modeling the parameter $\omega$ as a smooth function of the distance to the boundary and the angle between the animal's direction and the boundary normal vector. This makes the drift in the velocity equation dependent on the location process $X$ and the boundary of the domain.  \\ %More biblio about estimation methods

@@

Sound exposure is considered to affect only the parameters $\nu$ and $\tau$ in equation \ref{eq: RACVM equation}. This effect is estimated as a smooth function of a sound exposure covariate defined as the inverse of the distance to the ship. This choice of covariate is motivated by \cite{heide-jorgensen_behavioral_2021}. A response model is compared to a baseline model representing the narwhals movement before exposure to the disturbance, that is under normal conditions \cite{michelot_continuous-time_2022}.
The effect of sound exposure is assessed as a deviation in the response model parameters when compared to the baseline model parameters.

To summarize, the main contributions of this paper are:
\begin{itemize}
	\item Derivation of explicit formulas for the transition density of the "Rotational Advective Correlated Velocity Model" (RACVM, \ref{eq: RACVM equation}), and addition of this model in the framework of  the \texttt{R} package \texttt{smoothSDE}, to enable the use of smooth parameters depending on external covariates, and estimation from noisy observations irregularly spaced in time.
	\item Definition of a constrained version of the RCVM for simulation study, where deviations angles from shoreline and distance to shore are used as covariates to constrain the movement within a polygon, and align the velocity with the boundary of the domain.
	\item Application to real narwhal data to assess a behavioural response of the narwhals, using the exposure covariate defined as the inverse of the distance to the sound source.
\end{itemize}

@@

The land geometry for the specific region in South-East Greenland is gathered from \href{https://www.openstreetmap.org/#map=11/70.4029/-27.2928}{OpenStreetMap} database. The land polygons define the boundaries of the domain $\mathcal{D}$ in which the narwhals can move. The closest point on the shoreline to the observed GPS position $y_{ij}$  is denoted $p_{ij}$. Due to GPS measurement errors, the distance $D^{shore}_{ij}=d(y_{ij},p_{ij})$ is an approximation of the actual distance to the boundary. About $9\%$ of the GPS measurements in the dataset turned out to be on land. After visual examination, this was attributed to inaccuracies in the shoreline maps rather than errors in GPS measurements. We therefore used a satellite image of Sentinel-2 with 10 metres resolution downloaded from Google Earth Engine API to improve the accuracy of the land polygons. After adjustement of the polygon data using \texttt{qgis},  approximately $1 \%$ of the observations remained on land. Distance to shore values range from $0$ to $7.6$ km. We define an exposure to the shore covariate with the following formula
\begin{equation}
	\left\{
	\begin{array}{ll}
		E^{shore}_{ij}=0 & \mbox{if } D^{shore}_{ij}>D_{up} \\
		E^{shore}_{ij}=\frac{1}{D^{shore}_{ij}} & \mbox{if } D_{low}<D^{shore}_{ij} \leq D_{up} \\
		E^{shore}_{ij}=\frac{1}{D_{low}} & \mbox{ if } D^{shore}_{ij} \leq D_{low}
	\end{array}
	\right.
	\label{eq: exp shore definition}
\end{equation}
where the reference thresholds are $D_{low}=73$ m, which is the $5\%$ quantile of observed distances, and $D_{up}=830$ m, which is the median of observed distances. Thus we only consider the effect of the shore when the narwhals are close enough to it. 
The angle between the narwhal's heading and the shoreline $\Theta_{ij}$ is estimated as the angle between the observed empirical velocity $\hat{v}_{ij}=\frac{y_{i,j+1}-y_{ij}}{\Delta_j}$, where $\Delta_{ij}=t_{i,j+1}-t_{ij}$, and the vector $\vec{n}_{ij}=y_{ij}-p_{ij}$. This is illustrated in figure~\ref{fig: illustration nearest shore points}. A value $\Theta_{ij}=\pm \frac{\pi}{2}$ indicates movement parallel to the shore, while $\Theta_{ij} \in \left]\frac{\pi}{2},\pi\right] \cup \left[-\pi,-\frac{\pi}{2}\right[$ indicates movement toward the shore.

@@

We write $A=\begin{pmatrix} 
	\frac{1}{\tau} & -\omega \\
	\omega & \frac{1}{\tau}
\end{pmatrix}$ and recall equation~\ref{eq: RACVM equation}

\begin{equation*} \left\{
	\begin{array}{l}
		dX(t)=V(t)dt \\
		dV(t)=-A(V(t)-\mu)dt+\frac{2\nu}{\sqrt{\pi \tau}} dW(t) 
	\end{array}
	\right.
	\label{eq: RACVM equation bis}
\end{equation*}
where $X$ is the two-dimensional location and $V$ is the two-dimensional velocity.
In this section, we obtain explicit distributional results about the process $X$. Though this model is considered in several papers, we found no mention of the closed form formulas derived here. In \cite{gurarie_correlated_2017}, only the process $V$ is exhaustively studied, and the classical distributional results about this process are used for estimation and simulation purposes.
A more general formulation is considered by \cite{albertsen_generalizing_2018} in which the diffusion matrix is lower triangular matrix with positive diagonal elements whereas and two distinct autocorrelation parameters are considered in both directions allowing for anisotropic movement.
In this case, \cite{albertsen_generalizing_2018} derives an explicit form for the gaussian transition density of the velocity process, where the covariance matrix is expressed using a Kronecker sum \cite{albertsen_generalizing_2018}. But again, exact formulas for the process $X$ are not mentioned and the Euler-scheme is used to get approximate transition densities of $X$. Here, we claim that under the hypothesis of isotropic movement, with diagonal diffusion matrix as in equation ~\ref{eq: RACVM equation}, not only the velocity but also the position process have a closed form formula. \\

@@

Equation~\ref{eq: covariance matrix} shows that the two components of the location and velocity processes are independent.
Note that $\exp(-A\Delta)=e^{\frac{-\Delta}{\tau}} \begin{pmatrix} \cos(\omega \Delta) & \sin(\omega \Delta) \\ -\sin(\omega \Delta) & \cos(\omega \Delta) \end{pmatrix}$ is a weighted rotation matrix of angle $-\omega \Delta$.
Intuitively, ~\ref{eq: X V distribution} means that the next velocity $V(t+\Delta)$ is a weighted mean of the long term mean velocity $\mu$ and the previous velocity $V(t)$ rotated by an angle $-\omega \Delta$.
 Figure~\ref{fig: sample RACVM} shows one sample of a RCVM.\\

@@

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.5]{images/rcvm/illustrative_sample_standard.png}
	\caption{Trajectory of a RCVM with parameters $\tau=1$h, $\nu=4$ km/h, $\omega=0.5$ rad/h, $\mu=(0,0)$ km/h and constant time step $\Delta=5$ min. The red cross indicates the starting point of the process. The samples were obtained from the exact transition density as derived in ~\ref{prop: transition density}.}
	\label{fig: sample RACVM}
\end{figure}

Unconstrained movement of marine mammals  is often modeled through a CVM, that is equation~\ref{eq: RACVM equation} with $\omega=0$, or a RACVM with low angular velocity $\omega$.
In the case of motion under spatial constraints, we claim that an increase in tortuosity is the symptom of movement to avoid the boundary and adapt the heading to the spatial constraint.

@@

We thus propose to make use of a new RCVM that relies on the tortuosity parameter, or angular velocity $\omega$, to include how the animals turn in reaction to the shore. 
To do so, we introduce $\omega$ as a smooth function of the distance to the shore and the angle between the velocity vector and the shore normal vector. 
The smooth function should be non zero close to the shore, that is when a change of direction due to the proximity of the shore is expected, and  zero far away from the shore. 
We keep the notation $\mathcal{D} \subset \R^2$ for the polygon defining the domain of the process $X$. The new formulation of the equation is 
\begin{equation} \left\{
	\begin{array}{l}
		dX(t)=V(t) dt \\
		dV(t)=-A(X(t),V(t))V(t)dt+\frac{2\nu}{\sqrt{\pi \tau}} dW(t)

@@

with 
\begin{equation} 
		A(X(t),V(t))=\begin{pmatrix} 
			\frac{1}{\tau} & -\omega(X(t),V(t)) \\
			\omega(X(t),V(t)) & \frac{1}{\tau}
		\end{pmatrix}
	\label{eq: CRCVM matrix A}
\end{equation}
where $\omega(X(t),V(t))=f_{\omega}(D^{shore}(t),\Theta(t))$, $D^{shore}(t)$ is the observed distance to the boundary of the domain $\partial \mathcal{D}$
and $\Theta(t)$ is the observed angle between the velocity and the shoreline as defined in subsection ~\ref{subsection: covariates}. Thus, $f_{\omega}$ is a smooth function from $\R^{+} \times [-\pi,\pi]$ into $\R$.

Solving explicitly such a model is out of reach due to the non-linearity induced by the distance to the shore and the angle $\Theta$, but it is possible to get an approximate solution.
For any $t \geq 0$ and $\Delta>0$, if $\Delta$ is small enough, we can use the approximations
\[\forall s \in [t,t+\Delta], D^{shore}(s) \simeq D^{shore}(t) \quad \mbox{and} \quad  \Theta(s)\simeq \Theta(t)\]
In this case, we can use the formulas of section~\ref{section: RACVM} to get approximate transition densities.
The question still hanging is: what are the conditions on the function $f_{\omega}$ to ensure that the process is constrained ?
We don't formulate a comprehensive answer here, but only give examples of interpretable smooth functions that result in constrained samples of $X$.
An important spatial scale of the rotational movement is $\rho=\frac{\nu}{\omega}$, which is the diameter of rotation. It should always remain significantly lower than $D^{shore}$ for the process to be constrained. Moreover, a rotation needs to be considered near the boundary of the domain, that is when $D^{shore}$ approaches $0$, and when the animal is heading in the direction of the shore, that is for $\Theta \in \left[\frac{\pi}{2},\pi\right] \cup \left[-\pi,-\frac{\pi}{2}\right]$.
The closer we are to the shore, the faster the velocity should rotate,
whereas at a certain distance from the coast, there should be no further effect of the shore, meaning that $\omega$ should be close to $0$. Far from the coast, the movement model is simply an integrated Ornstein-Uhlenbeck process, as in the unconstrained case.
A weight based on the distance to the shore can be defined as follows :
\[c(D^{shore})=\exp\left(-\kappa \times \left(\frac{D^{shore}}{D_0}\right)^2\right)\mbox{ for some } D_0>0 \mbox{ and } \kappa>0\]

@@

This parametrization of $\omega$ is only motivated by the observations above, and do not pretend to be optimal. Any smooth function that meets minimum requirements could be used instead. An alternative is to specify the smooth function $f_{\omega}$ using spline basis functions. Figures~\ref{fig: smoothomega1} and~\ref{fig: smoothomega2} show a smooth function obtained from equation~\ref{eq: smooth omega}, while figures~\ref{fig: smoothomegasplines1} and~\ref{fig: smoothomegasplines2}  represents a smooth function $f_{\omega}$ obtained as a combination of nine cubic spline basis functions. In the sequel, such models with $\omega$ as a smooth function of $D^{shore}$ and $\Theta$ meeting minimum requirements will be refered to as Constrained Rotational Correlated Velocity Models (CRCVM).

@@

	\begin{subfigure}{0.48\textwidth}
	\centering
	\includegraphics[scale=0.3]{images/crcvm/smooth_splines_omega1.png}
	\caption{}
	\label{fig: smoothomegasplines1}
	\end{subfigure}
	\begin{subfigure}{0.48\textwidth}
	\centering
	\includegraphics[scale=0.3]{images/crcvm/smooth_splines_omega2.png}
	\caption{}
	\label{fig: smoothomegasplines2}
	\end{subfigure}
	\caption{Examples of smooth functions $f_{\omega}$. Angular velocity $\omega$ increases as $\Theta$ approaches $\pm\pi$, and decreases with increasing distance to the shore  (a)-(b) Two views of the smooth function $f_{\omega}$ defined by equation~\ref{eq: smooth omega} with values $D_0=0.3$ km, $\kappa=0.2$, $\lambda=2$, $\omega_0=\frac{\pi}{2}$ rad/min. (c)-(d) Two views of a smooth function $f_{\omega}$ obtained by combining nine bivariate cubic spline basis functions.}

@@

 This general model is very flexible, and different species-specific behaviours can be described. The function $\omega$ allows to control at which distance the animal will turn away from the shore, and how fast it will turn. We will show in section~\ref{section: simulation constrained motion} that is can be estimated from the observed positions, observed distances and angles. Specific constraints on the smooth function $f_{\omega}$ could also be set by biological expertise and knowledge about the species in question. Additionally, the tendency of an animal to move along the boundary of its spatial domain can be captured in the parameter $\tau$, as more persistent movement when the angle $\theta$ has value close to $\pm\frac{\pi}{2}$ would represent such an inclination. Again, this behavior can be statistically assessed from the observations.

 More precisely, to model persistent movement along the shoreline, we modify equation~\ref{eq: CRCVM matrix A} by expressing $\tau$ as a smooth function :
 \[\tau(X(t),V(t))=f_{\tau}(\Theta(t))\] where $f_{\tau}$ is a smooth function from $[-\pi,\pi]$ to $\R$.
 The function $f_{\tau}$ is defined such that it reaches local maximas in $\Theta=\pm\frac{\pi}{2}$, which represents movement along the boundary of the domain. An example of such a function is given in figure~\ref{fig:smoothtaubump}.
 \begin{figure}[H]
 	\centering
 	\includegraphics[scale=0.3]{images/crcvm/smooth_tau_bump}
 	\caption{Example of smooth function $f_{\tau}$ that can represent persistent movement along the boundary of the domain.}
 	\label{fig:smoothtaubump}
 \end{figure}

 We illustrate the flexibility of the CRCVMs in Figure~\ref{fig: crcvm examples}. Trajectories obtained within circular, rectangular and fjords domains for different smooth functions $f_{\omega}$ and $f_{\tau}$ are shown side by side.

@@

   	\hspace{2cm} 
 	 \begin{subfigure}{0.48\textwidth}
 	 	\includegraphics[scale=0.45]{images/crcvm/illustrative_sample_circ}
 	 	\label{fig:illustrativesamplecirc}
 	 \end{subfigure}

>>>>

\begin{abstract}
Narwhals in Arctic seas are increasingly exposed to human activities that can temporarily or permanently threaten their survival by modifying their behavior. We collected GPS data from a population of narwhals in the Scoresby Sound fjord system in Southeast Greenland, which were exposed to ships and seismic airguns during a controlled experiment in 2018. We propose a new continuous-time correlated velocity model with varying coefficients that includes spatial constraints on movement and sound exposure effects through covariates. Our model, applied to the narwhal data, demonstrates that sound exposure can disturb narwhals motion up to tens of kilometers. Specifically, we found a $40\%$ and $25\%$ increase in velocity and a $50\%$ and $30\%$ drop in persistence at distances of 4 km and 10 km from the sound source respectively.
\end{abstract}

@@

We focus on continuous-time models defined as a solution of a stochastic differential equation with time-varying coefficients. One widely known model for such data is the continuous-time correlated random walk, also known as the integrated Ornstein-Uhlenbeck or correlated velocity model (CVM) \cite{johnson_continuoustime_2008}. It can be viewed as a special case of velocity potential model, where the potential is a second degree polynomial \cite{preisler_analyzing_2013}. The following formulation of the equations represents an extension of the CVM which involves a rotational component and  is based on biologically interpretable parameters \cite{gurarie_correlated_2017}.
\begin{equation} \left\{
	\begin{array}{l}
		dX(t)=V(t)dt \\
		dV(t)=-\begin{pmatrix} 
			\frac{1}{\tau} & -\omega \\
			\omega & \frac{1}{\tau}
		\end{pmatrix}(V(t)-\mu)dt+\frac{2\nu}{\sqrt{\pi \tau}} dW(t) 
	\end{array}
	\right.
	\label{eq: RACVM equation}
\end{equation}
where $V(t)$ is the horizontal two-dimensional velocity at time $t$, $X(t)$ is the position, typically in longitude and latitude or in UTM coordinates and $W(t)$ is a two-dimensional brownian motion. 
The parameter $\tau$ is an autocorrelation time scale, $\nu$ controls the norm of the velocity and drives the random variability, $\mu$ is the long term velocity, and $\omega$ is an angular velocity that controls how fast the velocity vector rotates. The case $\omega=0$ corresponds to a standard CVM. In most applications, this standard model is satisfactory, but in case of tortuous movement, a non-zero value of $\omega$ is sometimes needed \cite{gurarie_correlated_2017,alt_correlation_1990,albertsen_generalizing_2018}. Moreover, whether or not to include a non-zero mean velocity parameter $\mu$ depends on the specific context of the study. Typically, it would make sense to have $\mu\neq (0,0)$ when examining migratory patterns or avoidance reactions from a fixed sound source. We follow \cite{gurarie_correlated_2017} and refer to~\eqref{eq: RACVM equation} as "Rotational Correlated Velocity Model" (RCVM) in the case $\mu=(0,0)$, and "Rotational Advective Correlated Velocity Model" (RACVM) in the case $\mu \neq (0,0)$.\\

In general, the motion of marine mammals is subject to spatial constraints induced by the coastline. This is particularly the case for the narwhal data we want to analyse: the animals move in a restricted domain that consists in a complex system of fjords in Scoresby Sound (South-East Greenland). Thus, constraints for the shore are needed in the location process $X$. However, this is not the case in model~\eqref{eq: RACVM equation}, nor in most of the SDE used for animal movement \cite{johnson_continuoustime_2008,michelot_varying-coefficient_2021,gurarie_correlated_2017}. 
Recently, a reflected version of the CVM has been considered to study sea lion telemetry data \cite{hanks_reflected_2017}.
In this study, simulation of constrained paths is achieved by projecting on the coastline the  points that eventually reach land. Estimation of the SDE parameters then relies on computationally intensive MCMC algorithm (Metroplis-Hastings), and requires high frequency data. Another idea is to add a repulsive potential function in the drift, which is done to constrain ants movement within a box in \cite{russell_spatially_2018}. In our case study, the narwhals move in a complex fjords system, which means that this technique might not be easily replicated because it relies on using very basic boundaries to derive an explicit expression for the potential function. As a consequence, there is a need to develop new models adapted to such spatial constraints.\\
Additionally, the way to model the motion of an animal near the boundary of the domain should be species-specific, accounting for the particular behaviors and spatial use of the studied animal \cite{brillinger_simulating_2003}. For narwhals, it was observed from the data that they have a proclivity to move along the shoreline. Here we propose a new model based on the following idea: as the narwhals approach the shore, at some point they need to rotate to avoid or align with the shoreline. The novelty of this paper consists in modeling the parameter $\omega$ as a smooth function of the distance to the boundary and the angle between the animal's direction and the boundary normal vector. This makes the drift in the velocity equation dependent on the location process $X$ and the boundary of the domain.  \\ %More biblio about estimation methods

@@

Sound exposure is considered to affect only the parameters $\nu$ and $\tau$ in equation \eqref{eq: RACVM equation}. This effect is estimated as a smooth function of a sound exposure covariate defined as the inverse of the distance to the ship. This choice of covariate is motivated by \cite{heide-jorgensen_behavioral_2021}. A response model is compared to a baseline model representing the narwhals movement before exposure to the disturbance, that is under normal conditions \cite{michelot_continuous-time_2022}.
The effect of sound exposure is assessed as a deviation in the response model parameters when compared to the baseline model parameters.

To summarize, the main contributions of this paper are:
\begin{itemize}
	\item Derivation of explicit formulas for the transition density of the "Rotational Advective Correlated Velocity Model" (RACVM, \eqref{eq: RACVM equation}), and addition of this model in the framework of  the \texttt{R} package \texttt{smoothSDE}, to enable the use of smooth parameters depending on external covariates, and estimation from noisy observations irregularly spaced in time.
	\item Definition of a constrained version of the RCVM for simulation study, where deviations angles from shoreline and distance to shore are used as covariates to constrain the movement within a polygon, and align the velocity with the boundary of the domain.
	\item Application to real narwhal data to assess a behavioural response of the narwhals, using the exposure covariate defined as the inverse of the distance to the sound source.
\end{itemize}

@@

The land geometry for the specific region in South-East Greenland is gathered from \href{https://www.openstreetmap.org/#map=11/70.4029/-27.2928}{OpenStreetMap} database. The land polygons define the boundaries of the domain $\mathcal{D}$ in which the narwhals can move. The closest point on the shoreline to the observed GPS position $y_{ij}$  is denoted $p_{ij}$. Due to GPS measurement errors, the distance $D^{shore}_{ij}=d(y_{ij},p_{ij})$ is an approximation of the actual distance to the boundary. About $9\%$ of the GPS measurements in the dataset turned out to be on land. After visual examination, this was attributed to inaccuracies in the shoreline maps rather than errors in GPS measurements. We therefore used a satellite image of Sentinel-2 with 10 metres resolution downloaded from Google Earth Engine API to improve the accuracy of the land polygons. After adjustement of the polygon data using \texttt{qgis},  approximately $1 \%$ of the observations remain on land. Distance to shore values range from $0$ to $7.6$ km. We define an exposure to the shore covariate with the following formula
\begin{equation}
	\left\{
	\begin{array}{ll}
		E^{shore}_{ij}=0 & \mbox{if } D^{shore}_{ij}>D_{up} \\
		E^{shore}_{ij}=\frac{1}{D^{shore}_{ij}} & \mbox{if } D_{low}<D^{shore}_{ij} \leq D_{up} \\
		E^{shore}_{ij}=\frac{1}{D_{low}} & \mbox{ if } D^{shore}_{ij} \leq D_{low}
	\end{array}
	\right.
	\label{eq: exp shore definition}
\end{equation}
where the reference thresholds are $D_{low}=73$ m, which is the $5\%$ quantile of observed distances, and $D_{up}=3$ km. Thus we only consider the effect of the shore when the narwhals are within a $3$ km range from the shore. 
The angle between the narwhal's heading and the shoreline $\Theta_{ij}$ is estimated as the angle between the observed empirical velocity $\hat{v}_{ij}=\frac{y_{i,j+1}-y_{ij}}{\Delta_j}$, where $\Delta_{ij}=t_{i,j+1}-t_{ij}$, and the vector $\vec{n}_{ij}=y_{ij}-p_{ij}$. This is illustrated in figure~\ref{fig: illustration nearest shore points}. A value $\Theta_{ij}=\pm \frac{\pi}{2}$ indicates movement parallel to the shore, while $\Theta_{ij} \in \left]\frac{\pi}{2},\pi\right] \cup \left[-\pi,-\frac{\pi}{2}\right[$ indicates movement toward the shore.

@@

We denote $A=\begin{pmatrix} 
	\frac{1}{\tau} & -\omega \\
	\omega & \frac{1}{\tau}
\end{pmatrix}$ and rewrite~\eqref{eq: RACVM equation}

\begin{equation*} \left\{
	\begin{array}{l}
		dX(t)=V(t)dt \\
		dV(t)=-A(V(t)-\mu)dt+\frac{2\nu}{\sqrt{\pi \tau}} dW(t) 
	\end{array}
	\right.
	\label{eq: RACVM equation bis}
\end{equation*}
In this section, we obtain explicit transition density for the process $U=\begin{pmatrix} X & V \end{pmatrix}^\top$. Though this model is considered in several papers, we found no mention of the closed form formulas derived here. In \cite{gurarie_correlated_2017}, only the process $V$ is exhaustively studied, and the classical distributional results about this process are used for estimation and simulation purposes.
A more general formulation is considered by \cite{albertsen_generalizing_2018} in which the diffusion matrix is lower triangular matrix with positive diagonal elements whereas and two distinct autocorrelation parameters are considered in both directions allowing for anisotropic movement.
In this case, \cite{albertsen_generalizing_2018} derives an explicit form for the gaussian transition density of the velocity process, where the covariance matrix is expressed using a Kronecker sum \cite{albertsen_generalizing_2018}. But again, exact formulas for the process $X$ are not mentioned and the Euler-scheme is used to get approximate transition densities of $X$. Here, we claim that under the hypothesis of isotropic movement, with diagonal diffusion matrix as in equation ~\eqref{eq: RACVM equation}, not only the velocity but also the position process have a closed form formula. \\

@@

Equation~\eqref{eq: covariance matrix} shows that the two components of the location and velocity processes are independent.
Note that $\exp(-A\Delta)=e^{\frac{-\Delta}{\tau}} \begin{pmatrix} \cos(\omega \Delta) & \sin(\omega \Delta) \\ -\sin(\omega \Delta) & \cos(\omega \Delta) \end{pmatrix}$ is a weighted rotation matrix of angle $-\omega \Delta$.
Intuitively, ~\eqref{eq: X V distribution} means that the next velocity $V(t+\Delta)$ is a weighted mean of the long term mean velocity $\mu$ and the previous velocity $V(t)$ rotated by an angle $-\omega \Delta$.
 Figure~\ref{fig: sample RACVM} shows one sample of a RCVM.\\

@@

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.3]{images/rcvm/illustrative_sample_standard.png}
	\caption{Trajectory of a RCVM with parameters $\tau=1$h, $\nu=4$ km/h, $\omega=0.5$ rad/h, $\mu=(0,0)$ km/h and constant time step $\Delta=5$ min. Values of $x$ and $y$ are in km. The red cross indicates the starting point of the process. The samples were obtained from the exact transition density as derived in ~\ref{prop: transition density}.}
	\label{fig: sample RACVM}
\end{figure}

Unconstrained movement of marine mammals  is often modeled through a CVM, that is~\eqref{eq: RACVM equation} with $\omega=0$, or a RACVM with low angular velocity $\omega$.
In the case of motion under spatial constraints, we claim that an increase in tortuosity is the symptom of movement to avoid the boundary and adapt the heading to the spatial constraint.

@@

We thus propose a new RCVM that relies on the tortuosity parameter, or angular velocity $\omega$, to describe how the animals turn in reaction to the shore. 
To do so, we consider $\omega$ as a smooth function of the distance to the shore and the angle between the velocity vector and the shore normal vector. 
The smooth function should be non zero close to the shore, that is when a change of direction due to the proximity of the shore is expected, and  zero far away from the shore. 
We keep the notation $\mathcal{D} \subset \R^2$ for the polygon defining the domain of the process $X$. The new formulation of the equation is 
\begin{equation} \left\{
	\begin{array}{l}
		dX(t)=V(t) dt \\
		dV(t)=-A(X(t),V(t))V(t)dt+\frac{2\nu}{\sqrt{\pi \tau}} dW(t)

@@

with 
\begin{equation} 
		A(X(t),V(t))=\begin{pmatrix} 
			\frac{1}{\tau} & -\omega(X(t),V(t)) \\
			\omega(X(t),V(t)) & \frac{1}{\tau}
		\end{pmatrix}
	\label{eq: CRCVM matrix A}
\end{equation}
where $\omega(X(t),V(t))=f_{\omega}(D^{shore}(t),\Theta(t))$ with  $f_{\omega}$ a smooth function from $\R^{+} \times [-\pi,\pi]$ into $\R$ meeting minimum requirements. In the sequel, such models will be refered to as Constrained Rotational Correlated Velocity Models (CRCVM).\\

Solving explicitly such a model is out of reach due to the non-linearity induced by the distance to the shore and the angle $\Theta$, but it is possible to get an approximate solution.
For any $t \geq 0$ and $\Delta>0$, if $\Delta$ is small enough, we can use the approximations
\[\forall s \in [t,t+\Delta], D^{shore}(s) \simeq D^{shore}(t) \quad \mbox{and} \quad  \Theta(s)\simeq \Theta(t)\]
In this case, we use the formulas of section~\ref{section: RACVM} to get approximate transition densities.
The question still hanging is: what are the conditions on the function $f_{\omega}$ to ensure that the process is constrained ?
We don't formulate a comprehensive answer here, but only give examples of interpretable smooth functions that result in constrained samples of $X$.
An important spatial scale of the rotational movement is $\rho=\frac{\nu}{\omega}$, which is the diameter of rotation. It should always remain significantly lower than $D^{shore}$ for the process to be constrained. Moreover, a rotation needs to be considered near the boundary of the domain, that is when $D^{shore}$ approaches $0$, and when the animal is heading in the direction of the shore, that is for $\Theta \in \left[\frac{\pi}{2},\pi\right] \cup \left[-\pi,-\frac{\pi}{2}\right]$.
The closer we are to the shore, the faster the velocity should rotate,
whereas at a certain distance from the coast, there should be no further effect of the shore, meaning that $\omega$ should be close to $0$. Far from the coast, the movement model is simply an integrated Ornstein-Uhlenbeck process, as in the unconstrained case.
A weight based on the distance to the shore can be defined as follows :
\[c(D^{shore})=\exp\left(-\kappa \times \left(\frac{D^{shore}}{D_0}\right)^2\right)\mbox{ for some } D_0>0 \mbox{ and } \kappa>0\]

@@

This parametrization of $\omega$ is only motivated by the observations above, and do not pretend to be optimal. Any smooth function that meets minimum requirements could be used instead. An alternative is to specify the smooth function $f_{\omega}$ using spline basis functions. Figures~\ref{fig: smoothomega1} and~\ref{fig: smoothomega2} show function~\ref{eq: smooth omega}, while Figures~\ref{fig: smoothomegasplines1} and~\ref{fig: smoothomegasplines2}  represent a smooth function $f_{\omega}$ obtained as a combination of nine cubic spline basis functions approximating~\ref{eq: smooth omega}. Sampled trajectories of these models are shown in Figure~\ref{fig: crcvm examples}.

@@

	\begin{subfigure}{0.48\textwidth}
	\centering
	\includegraphics[scale=0.3]{images/crcvm/smooth_splines_omega1.png}
	\caption{}
	\label{fig: smoothomegasplines1}
	\end{subfigure}
	\begin{subfigure}{0.48\textwidth}
	\centering
	\includegraphics[scale=0.3]{images/crcvm/smooth_splines_omega2.png}
	\caption{}
	\label{fig: smoothomegasplines2}
	\end{subfigure}
	\caption{Examples of smooth functions $f_{\omega}$. Angular velocity $\omega$ increases as $\Theta$ approaches $\pm\pi$, and decreases with increasing distance to the shore  (a)-(b) Two views of the smooth function $f_{\omega}$ defined by~\eqref{eq: smooth omega} with values $D_0=0.3$ km, $\kappa=0.2$, $\lambda=2$, $\omega_0=\frac{\pi}{2}$ rad/min. (c)-(d) Two views of a smooth function $f_{\omega}$ obtained by combining nine bivariate cubic spline basis functions.}

@@

 This general model is very flexible, and different species-specific behaviours can be described. The function $\omega$ allows to control at which distance the animal will turn away from the shore, and how fast it will turn. We will show in section~\ref{section: simulation constrained motion} that is can be estimated from the observed positions, distances and angles. Specific constraints on the smooth function $f_{\omega}$ could also be set by biological expertise and knowledge about the species in question. Additionally, the tendency of an animal to move along the boundary of its spatial domain can be captured in the parameter $\tau$, as more persistent movement when the angle $\theta$ has value close to $\pm\frac{\pi}{2}$ would represent such an inclination. More precisely, to model persistent movement along the shoreline, we modify~\eqref{eq: CRCVM matrix A} by expressing $\tau$ as a smooth function :
 \[\tau(X(t),V(t))=f_{\tau}(\Theta(t))\] with $f_{\tau}:[-\pi,\pi] \longrightarrow \R$.
 The function $f_{\tau}$ is defined such that it reaches local maximas in $\Theta=\pm\frac{\pi}{2}$, which represents movement along the boundary of the domain. An example of such a function is given in figure~\ref{fig:smoothtaubump}.
 \begin{figure}[h]
 	\centering
 	\includegraphics[scale=0.3]{images/crcvm/smooth_tau_bump}
 	\caption{Example of smooth function $f_{\tau}$ that can represent persistent movement along the boundary of the domain.}
 	\label{fig:smoothtaubump}
 \end{figure}

 The flexibility of CRCVMs is illustrated in Figure~\ref{fig: crcvm examples}. Trajectories obtained within circular, rectangular and fjords domains for different smooth functions $f_{\omega}$ and $f_{\tau}$ are shown side by side.

@@

   	\hspace{3cm} 
 	 \begin{subfigure}{0.48\textwidth}
 	 	\includegraphics[scale=0.45]{images/crcvm/illustrative_sample_circ}
 	 	\label{fig:illustrativesamplecirc}
 	 \end{subfigure}

   	\hspace{3cm} 
  	 \begin{subfigure}{0.48\textwidth}
  		\includegraphics[scale=0.45]{images/crcvm/illustrative_sample_rect}
  		\label{fig:illustrativesamplerect}
 	 \end{subfigure}
